---
title: 'Practical 4: Deep Learning'
author: "Nina van Gerwen (1860852)"
date: "2022-12-08"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## 2: Take-home Exercises: a DFF NN

Here we will develop a *deep feed-forward* neural network! Fun times await.

### 2.1: Data preparation

First, we prepare the data and load the proper packages.

```{r}
library(tidyverse)
library(keras)
```


#### 2.1.1: Loading the MNIST dataset

```{r}
mnist <- dataset_mnist()
```


#### 2.1.2: Creating a plotting function

```{r}
plot_img <- function(img, col = gray.colors(255, start = 1, end = 0), ...) {
  image(t(img), asp = 1, ylim = c(1.1, -0.1), col = col, bty = "n", axes = FALSE, ...)
}
```

Code was taken from the practical. 

#### 2.1.3: Ensuring brightness values of the images are in the range (0, 1)

```{r}
range(mnist$train$x)
mnist$train$x <- mnist$train$x / 255
range(mnist$train$x)
mnist$test$x <- mnist$test$x / 255
```

To ensure this, we look at the range of the brightness values, we find that
they range from 0 to 255. Therefore, to keep it between 0 and 1, we simply divide
all brightness values by 255. Then we find that the range of brightness values
is between 0 and 1. 

### 2.2: Multinomial logistic regression

```{r}
set.seed(1248)

multinom  <- 
  keras_model_sequential(input_shape = c(28, 28)) %>% # initialize a sequential model
  layer_flatten() %>% # flatten 28*28 matrix into single vector
  layer_dense(10, activation = "softmax") # softmax outcome == logistic regression for each of 10 outputs

multinom$compile(
  loss = "sparse_categorical_crossentropy", # loss function for multinomial outcome
  optimizer = "adam", # we use this optimizer because it works well
  metrics = list("accuracy") # we want to know training accuracy in the end
)

summary(multinom)
```

#### 2.2.4: Explaining the number of parameters

The images have 28 x 28 pixels. This means that they have 28 x 28 dimensions.
This leads to a total of 784 parameters/weights per outcome variable for only the features. Then we also have to add another one for the intercept/bias per outcome variable, leading to 785. Finally, considering that there are 10 outcome variables, and all pixels can load on every outcome variable with their own intercept, it means that there are 785 x 10 = 7850 parameters in total!

#### 2.2.5: Accuracy for a multinomial logistic regression model that was trained for 5 epochs

```{r}
multinom %>% fit(x = mnist$train$x, y = mnist$train$y, epochs = 5, validation_split = 0.2, verbose = 1)
```

From the results, we find that after training for 5 epochs, the multinomial
logistic regression has a validation accuracy of approximately 0.925.

#### 2.2.6: Accuracy for the multinomial logistic regression after another 5 epochs

```{r}
multinom %>% fit(x = mnist$train$x, y = mnist$train$y, epochs = 5, validation_split = 0.2, verbose = 1)
```

After training the model for another 5 epochs, the validation accuracy has
risen to 0.929. We are probably simply overfitting the model, however.

### 2.3: Deep Feed-Forward Neural Network

#### 2.3.7: Creating our own DFF NN with 50890 parameters and training it for 10 epochs

```{r}
dff_nn <- 
  ## first, we tell the model what shape the input is
  keras_model_sequential(input_shape = c(28, 28)) %>% 
  ## we tell it to flatten the input to a vector
  layer_flatten() %>%
  ## we tell it we want a dense layer with 64 hidden units and relu activation function
  layer_dense(64, activation = "relu") %>%
  ## then we want an outcome layer (also dense) with the number of units
  ## equal to the number of dummy categories and softmax activation function
  layer_dense(10, activation = "softmax")

## Then we change the compiling to:
dff_nn$compile(
  loss = "sparse_categorical_crossentropy", # the same loss function as before
  optimizer = "adam", # we use this optimizer because it works well
  metrics = list("accuracy") # we want to know training accuracy in the end
)

## And finally, we train the model for 10 epochs
dff_nn %>% fit(x = mnist$train$x, y = mnist$train$y, epochs = 10, validation_split = 0.2, verbose = 1)
```

After 10 epochs, the DFF NN model has a validation accuracy of approximately 0.972.

#### 2.2.8: Creating predictions for the test data using the two models and comparing performance

```{r}
class_predict <- function(model, x_train) predict(model, x = x_train) %>% apply(1, which.max) - 1

multinom_pred <- class_predict(multinom, x_train = mnist$test$x)

dff_nn_pred <- class_predict(dff_nn, x_train = mnist$test$x)

## For multinomial logistic regression:
table(true = mnist$test$y, pred = multinom_pred)

## For DFF NN
table(true = mnist$test$y, pred = dff_nn_pred)
```


In the multinomial logistic regression, we find more wrongly categorised numbers
(as you would expect). In both cases, the number 4 is very often predicted as a 9. However, less
in the deep feed-forward neural network. In multinomial, a 5 is olso often categorised
as either a 2 or 8. This happens less in the deep feed-forward neural network.

#### 2.2.9: Another deep feed-forward neural network

```{r}
dff_nn_2 <- 
  keras_model_sequential(input_shape = c(28, 28)) %>% 
  layer_flatten() %>%
  ## the only extra layer we add is another dense layer with 128 hidden units
  layer_dense(128, activation = "relu") %>%
  layer_dense(64, activation = "relu") %>%
  layer_dense(10, activation = "softmax")

dff_nn_2$compile(
  loss = "sparse_categorical_crossentropy",
  optimizer = "adam",
  metrics = list("accuracy") 
)

## And finally, we train the model for 10 epochs
dff_nn_2 %>% fit(x = mnist$train$x, y = mnist$train$y, epochs = 10, validation_split = 0.2, verbose = 1)

dff_nn_2_pred <- class_predict(dff_nn_2, x_train = mnist$test$x)

table(true = mnist$test$y, pred = dff_nn_2_pred)
```

The validation accuracy for this second DFF NN is 0.972, similar to the other
DFF NN. However, the confusion matrix does show some differences. Largest difference
is the amount of missclassifications for the number 5, which is often
classified as a 3. This did not happen as often in the other models. Furthermore,
this model has a lot more parameters. For parsimonity, I would choose the other
DFF NN.

### 3: Lab Exercises

### 3.1: Convolutional Neural Network

#### 3.1.10: 





